---
title: "Supermarket Sales Time Series Investigation: Part 1"
output: rmarkdown::github_document
always_allow_html: yes
---

### Notebook Context
- time series analysis using [supermarket sales data from Kaggle](https://www.kaggle.com/c/store-sales-time-series-forecasting/data)
- work with data at monthly aggregated level to start given common business scenario to analyze monthly sales/trends
- techniques and method inspired by [Forecasting Principles and Practice | Hyndman & Athanasopoulos](https://otexts.com/fpp3/)

### Objective
1. Visualize sales time series trends
2. Decompose monthly sales into trend-cycle and seasonal components
3. Identify product families that are trending and/or seasonal
4. Apply simple methods for forecast monthly average sales per day

```{r echo=FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(fpp3)
library(tidyverse)
library(ggfortify)
library(gridExtra)
```

### Data Prep 
```{r}
### use Kaggle training dataset to represent our "business data"
train_raw <- read_csv("store-sales-time-series-forecasting/train.csv")
```

#### Monthly Sales
- aggregate to monthly sales by product family across stores
- use tsibble object for time series analysis
```{r}
monthly_sales_ts <- train_raw %>%
      ### aug is a partial month so we filter to dates prior to Aug 1
      filter(date < '2017-08-01') %>%
      group_by(family, 
               year_month = yearmonth(format_ISO8601(date, precision ="ym"))) %>%
      summarise(total_sales = sum(sales),
                sales_days = n_distinct(date),
                avg_sales_per_day = total_sales/sales_days) %>%
      as_tsibble(index = year_month,
                 key = c("family"))

head(monthly_sales_ts)
```

### 1. Visualize sales time series trends
#### Sales by Year
- Jan - Jul Sales increase YoY
- Visualize Jan - Jul sales by year due to most recent year being a partial year
```{r}
monthly_sales_ts %>%
   filter(month(year_month)<=7) %>%
   ungroup() %>%
   index_by(year = year(year_month)) %>%
   summarise(total_sales = sum(total_sales)) %>%
   autoplot(total_sales) +
  labs(title="Sales trend by year",
    y="Total Sales",
    x="Year")
```

#### Sales by Month
- sales data from stores in Ecuador (macro country context could likely shed light on some of the large dips)
- for example, late 2014/early 2015 seems to correlate with oil price crash
- large part of Ecuador's economy is based on the export of oil
- callout: calendar day bias here (not all calendar days have equal number of sales days)
- log transformed variable attempts to stabilize variance
```{r}
monthly_sales_ts %>%
  ungroup() %>%
  index_by() %>%
  summarise(
    total_sales = sum(total_sales)
  ) %>%
  autoplot(vars(total_sales, log10(total_sales))) +
  labs(title="Monthly sales trend by year month",
       y="See panel header",
       x="Year Month")
```

#### Aggregated monthly sales data
- generate monthly sales time series tsibble object
- "avg_sales_per_day" used to control for number of sales days difference by calendar month
```{r}
cal_month_sales <- monthly_sales_ts %>%
   ungroup() %>%
   index_by(year_month) %>%
   summarise(avg_sales_per_day = sum(total_sales)/unique(sales_days)) %>%
   as_tsibble(index=year_month, key=NULL)
```

#### Average sales per day trend
- used to control for number of sales days difference by calendar month
```{r}
cal_month_sales %>%
   index_by(year(year_month)) %>%
   mutate(avg_avg_sales_per_day_per_year = mean(avg_sales_per_day)) %>%
   autoplot(avg_sales_per_day) +
   geom_line(aes(y=avg_avg_sales_per_day_per_year, group=year(year_month), 
                 color="Avg. of (Avg. Sales per Day) per Year"),
             linetype="dashed",
             alpha=0.5,
             size=1.5) +
   theme(legend.position = "top") +
   labs(title="Avg. Sales per Day by Year Month",
        y="Avg. Sales per Day",
        x="Year Month",
        color="")
```

#### Compare avg sales per day YoY performance by month
```{r}
cal_month_sales %>%
      gg_subseries(avg_sales_per_day) +
      labs(title="Avg. Sales per Day by Year per Month",
        y="Avg. Sales per Day",
        x="Year",
        color="")
```

#### Avg. Sales per Day vs Scaled Avg. Sales per Day
- scaled "avg_sales_per_day" feature attempts to strip out increasing trend YoY and investigate with year seasonality 
- scaling by year we subtract the mean and divide by the standard deviation
- Dec is consistently well above the mean per pear
- Feb tends to be lowest sales per day month 
- regarding Feb, in practice, we'd want to investigate if country context might explain this or if there's still calendar bias after the average sales per day approach applied
- the scaling approach helps to control for different sales levels year to year

```{r}
p1_and_2_theme <- theme(
   axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

p1 <- cal_month_sales %>%
  gg_season(avg_sales_per_day, period = "year") +
  p1_and_2_theme +
  labs(subtitle="Avg. Sales per Day by Year Month",
       y="Avg. Sales per Day",
       x="Year Month")
     
p2 <- cal_month_sales %>%
  index_by(year = year(year_month)) %>%
  mutate(year_scaled_avg_sales_per_day = scale(avg_sales_per_day)) %>%
  gg_season(year_scaled_avg_sales_per_day, period = "year") +
  p1_and_2_theme +
  labs(subtitle="Avg. Sales per Day (scaled) by Year Month",
       y="Avg. Sales per Day (year scaled)",
       x="Year Month")

grid.arrange(p1, p2, ncol=1)
```

### 2. Decompose monthly sales into trend-cycle and seasonal components
- used to help improve understanding of a time series and sometimes used to improve forecast
- decomposition can also be used to generate seasonally adjusted data
- lecture notes from Professor 
- STL: seasonal and trend decomposition with loess

#### STL decomposition
```{r}
cal_month_sales %>%
   model(
      STL(avg_sales_per_day)
   ) %>%
   components() %>%
   autoplot()
```

#### STL decomposition remainder component
- ACF plot used to confirm there still seems to be signal in the remainder component
- if the remainder component resembled white noise we should see most data fall within bounds
- trend-cycle and seasonality components not capturing all signal

```{r}
cal_month_sales %>%
   model(
      STL(avg_sales_per_day)
   ) %>%
   components() %>%
   # features(remainder, ljung_box)
   ACF(remainder) %>%
   autoplot() +
   labs(title="ACF plot STL remainder component")
```

### 3. Identify product families that are trending and/or seasonal
- apply STL time series decomposition and generate feature measures to describe characteristics of each product family time series
- dairy, bread/bakery, deli are product family groups with highest trend strength
- frozen foods, alcohol, grocery 2 have the highest seasonal strength

Trending product areas
```{r}
stl_features <- monthly_sales_ts %>% 
   features(avg_sales_per_day, feat_stl)

monthly_sales_ts %>% 
   inner_join(stl_features) %>%
   mutate(family_trunc = str_trunc(family, 20)) %>%
   autoplot(avg_sales_per_day, color="dodgerblue") +
   facet_wrap(. ~ reorder(family_trunc, -trend_strength), nrow=5, scales="free") +
   theme(legend.position = "none",
         axis.text = element_blank(),
         axis.ticks = element_blank(),
         strip.text = element_text(size=4)) +
   labs(title="Product family with highest to lowest trend 
strength (left to right, top to bottom)",
        subtitle = "Sparkline trends",
        x="Year Month",
        y="Avg. Sales per Day")
```

Seasonal product areas
```{r}
monthly_sales_ts %>% 
   inner_join(stl_features) %>%
   mutate(family_trunc = str_trunc(family, 20)) %>%
   autoplot(avg_sales_per_day, color="forestgreen") +
   facet_wrap(. ~ reorder(family_trunc, -seasonal_strength_year), nrow=5, scales="free") +
   theme(legend.position = "none",
         axis.text = element_blank(),
         axis.ticks = element_blank(),
         strip.text = element_text(size=4)) +
   labs(title="Product family with highest to lowest 
yearly seasonal strength (left to right, top to bottom)",
        subtitle = "Sparkline trends",
        x="Year Month",
        y="Avg. Sales per Day")
```

### 4. Simple methods for forecast monthly sales (using previous monthly sales data)
- simple forecast methods often used as baseline for more complex methods
- simple methods often work well when series resembles white noise / random walk
- future investigation area: bottoms-up forecasting approach could forecast product family areas and/or generate forecasts by store + other projects explore more complex techniques

#### Fit several models
- fit model on 2013 to 2016 data
- model accuracy metrics lowest for tslm model
- forecast visualizations also confirm tslm gets closest to actual data
- tslm: time series linear regression which creates features for trend and seasonality
- mean: set forecast to training data average
- naive: set the forecast to the value of the last observation
- snaive: set the forecast to the last value for the same season of the prior year
- drift: drawing a line between the first and last points and extrapolating it

```{r}
train_fit <- cal_month_sales %>%
   filter_index("Jan 2013" ~ "Dec 2016") %>%
   model(tslm = TSLM(avg_sales_per_day ~ trend() + season()),
         mean = MEAN(avg_sales_per_day),
         naive = NAIVE(avg_sales_per_day),
         snaive = SNAIVE(avg_sales_per_day ~ lag("year")),
         drift = NAIVE(avg_sales_per_day ~ drift()))

accuracy(train_fit) %>%
   select(.model, RMSE, MAE, MAPE, MASE)
```
```{r}
forecast_7_months <- train_fit %>%
      forecast(h=7)

forecast_7_months %>%
   autoplot() +
   autolayer(cal_month_sales %>% filter_index("Jan 2016" ~ .)) +
   facet_grid(. ~ .model) +
   theme(legend.position = "top",
         axis.text.x = element_text(angle = 45, hjust = 1, 
                                    vjust = 1, size=6)) +
   scale_y_continuous(label=scales::comma) +
   labs(title="Forecast Jan to Jul 2017 Avg. Sales per Day",
        y="Avg. Sales per Day",
        x="Year Month",
        color="Model Forecast",
        fill="Model Forecast")
```

Part 2 to explore more robust forecasting techniques and bottoms-up forecasting.